---
title: "Custom Data Standardization (Undefined Technology)"
author: "scRNA-seq Pipeline"
format: 
  html:
    toc: true
    theme: flatly
execute:
  warning: false
---

## 1. Introduction & Setup

This document is a template for standardizing single-cell datasets from custom or undefined technologies. Your goal is to transform your raw data into a strictly formatted sparse matrix, a `Genes.txt` file, and a `Cells.csv` metadata file for each sample. The metadata information for each sample will be collected and saved in a dataset wide `Samples.txt`.

::: callout-important
### TODO: Define paths and names

Write the dataset specific paths and file names. If you know where the working directory leads to, than you can use relative paths, else absolute paths are recommended.
:::

```{r Path definitions}
rm(list = ls(all.names = TRUE))
gc()

# Fill out path string
app_folder_path <- "/path/to/app_folder"
dataset_dir_path <- "/path/to/dataset_folder"
resources_path <- "/path/to/resources_folder"
```

```{r setup}
#| message: false
library(data.table)
library(stringr)
library(Matrix)
library(yaml)

# 1. Load general config
config <- read_yaml(file.path(app_folder_path, 'general_config.yaml'))

# 2. Source utility scripts
invisible(lapply(list.files(file.path(app_folder_path, 'util_scripts'), pattern = '\\.R$', full.name = TRUE), source))

# 3. Setup paths and load dataset configs
processed_dataset_dir_path <- file.path(dataset_dir_path, config$dataset_paths$processed_dir)
raw_data_dir_path <- file.path(dataset_dir_path, config$dataset_paths$raw_dir)

dataset_config_path <- file.path(dataset_dir_path, config$config_files$dataset)
dataset_config <- read_yaml(dataset_config_path)

samples_config_path <- file.path(dataset_dir_path, config$config_files$samples)
samples_config <- read_yaml(samples_config_path)

# Load global resources (if necessary)

# Example of helpful resources
#hgnc_complete_set <- fread(file.path(resources_path, #config$resource_files$hgnc_complete_set), key = 'symbol')
#hgnc_complete_set <- hgnc_complete_set[!(ensembl_gene_id %in% #names(table(ensembl_gene_id))[table(ensembl_gene_id) > 1])]
```

## 2. Pre-validation

Before continuing, we verify that you have filled out the `samples_config.yaml` file for this workspace. If this fails, please open that file, fill in your sample metadata, and re-run.

```{r pre-validation}
# This will trigger a hard stop if samples_config.yaml is empty
prevalidate_setup(dataset_dir_path, config)

samples_info_dt <- rbindlist(samples_config$samples, fill = TRUE, use.names = TRUE)
sample_names <- samples_info_dt$id
```

## 3. Data Ingestion (USER ACTION REQUIRED)

::: callout-important
### TODO: Load Your Custom Data

Write your custom R code below to load your raw data. Because your data could be in CSVs, H5AD, RDS, or a proprietary format, you must manually read it into the R environment.
:::

```{r load-raw-data}
# --- USER INPUT REQUIRED ---
# Example: 
# raw_matrix <- read.csv("data/extracted/my_custom_counts.csv", row.names = 1)
# raw_metadata <- fread("data/extracted/my_custom_metadata.csv")

# Ensure your matrix is converted to a sparse format to save memory!
# expmat <- as(as.matrix(raw_matrix), "CsparseMatrix")
```

## 4. Matrix Inspection & Configuration

Depending on the source of your dataset, the expression matrix might be raw UMI counts, TPM, or log-normalized. Inspect the data below to decide how to name your final matrix file.

```{r matrix-inspection}
# Inspect the first few column sums and maximum values
# summary(Matrix::colSums(expmat))
# max(expmat)
```

::: callout-note
### Update Matrix State

Based on the output above, update the `matrix_state` variable. \* If values are strictly integers, it is likely `"UMIcounts"`. \* If column sums equal exactly 10k or 1M, it is likely `"CP10k"` or `"TPM"`.
:::

```{r update-config}
# --- USER INPUT REQUIRED ---
matrix_state <- "UMIcounts" # Change this if necessary (e.g., "TPM", "CP10k", "log2")

# Automatically construct and update the dataset configuration
new_expmat_name <- paste0("Exp_data_", matrix_state, ".mtx")
dataset_config$std_names$expmat <- new_expmat_name
write_yaml(dataset_config, dataset_config_path)

message("Updated dataset_config.yaml! Expected matrix name is now: ", new_expmat_name)
```

## 5. Standardization & Saving (USER ACTION REQUIRED)

::: callout-important
### TODO: Format and Save Data per Sample

You must split your data by sample (matching the IDs in `samples_config.yaml`). For each sample, you need to create and save: 1. `expmat`: A sparse matrix (`dgCMatrix`) with gene symbols as rownames and cell names as colnames. 2. `cells`: A data.table containing at minimum `cell_name`, `sample`, `patient`, and `complexity` (number of non-zero genes). 3. `genes`: A vector of mapped HGNC gene symbols.

Hint: Wrap the data.table assignments in `invisible()` to suppress Quarto printing
:::

```{r standardize-and-save}
# --- USER INPUT REQUIRED ---
sample_cell_counts <- list()

# Example Loop Skeleton:
# for (s_name in sample_names) {
#   message("Processing: ", s_name)
#   
#   1. Subset your matrix and metadata for this sample
#   sample_expmat <- expmat[, raw_metadata$sample == s_name]
#   
#   2. Ensure genes are HGNC symbols (Map from Ensembl if necessary using map_ensembl_to_symbol_vector)
#   
#   3. Build the Cells data.table (can use build_cells() function to get cells template table and fill out columns)
#   cells <- build_cells(colnames(sample_expmat))
#   
#   invisible(cells[, complexity := matkot::col_nnz(expmat)])
#   
#   4. Create Directory and Save Files
#   sample_dir <- file.path(processed_dataset_dir_path, s_name)
#   dir.create(sample_dir, showWarnings = FALSE)
#   
#   Matrix::writeMM(sample_expmat, file = file.path(sample_dir, dataset_config$std_names$expmat))
#   fwrite(cells, file = file.path(sample_dir, dataset_config$std_names$cells))
#   fwrite(data.table(rownames(sample_expmat)), file = file.path(sample_dir, dataset_config$std_names$genes), col.names = FALSE)
# }
```

## 6. Finalize Pipeline Configuration

This automated step creates the final `Samples.csv` registry and initializes the empty annotation cache required by the Shiny app.

```{r finalize-configs}
# Build Samples.csv 
samples <- build_samples(sample_names) # from utils

cols_to_add <- setdiff(names(samples_info_dt), "id")
samples[samples_info_dt, (cols_to_add) := mget(paste0("i.", cols_to_add)), on = .(sample = id)]

# Map cell counts (requires sample_cell_counts list to be populated in Step 5)
# invisible(samples[, n_cells := unlist(sample_cell_counts)[sample]])

fwrite(samples, file = file.path(processed_dataset_dir_path, dataset_config$std_names$samples))

# Setup selected_config.yaml based on samples
selected_config <- setup_selected_config(samples$sample)
write_yaml(selected_config, file.path(dataset_dir_path, config$config_files$selected))
```

## 7. Post-Validation

Run this final check to ensure the output matches the strict structural requirements of the Shiny App.

```{r post-validation}
#| output: asis

# This will generate HTML warning boxes if something is wrong, 
# or a green success box if you are ready to proceed!
validate_standardization(
  processed_dir_path = processed_dataset_dir_path, 
  sample_names = sample_names, 
  std_names = dataset_config$std_names
)
```
