---
title: "Data Standardization (10x Technology)"
author: "scRNA-seq Pipeline"
format: 
  html:
    toc: true
    theme: flatly
execute:
  warning: false
---

## 1. Introduction & Setup

This document standardizes raw 10x Genomics single-cell datasets. It extracts the raw archive, maps the files (matrix, features, barcodes) using naming conventions, allows for manual sample ID alignment, and strictly formats the output for the downstream pipeline.

::: callout-important
### TODO: Define paths and names

Write the dataset specific paths and file names. If you know where the working directory leads to, than you can use relative paths, else absolute paths are recommended.
:::

```{r Path definitions}
rm(list = ls(all.names = TRUE))
gc()

# Fill out path string
app_folder_path <- "/path/to/app_folder"
dataset_dir_path <- "/path/to/dataset_folder"
resources_path <- "/path/to/resources_folder"

# Fill out when data saved as archive and needs to be extracted
extract_archive <- FALSE
archive_name <- "<file_name>.tar"

```

```{r setup}
#| message: false
library(data.table)
library(stringr)
library(Matrix)
library(yaml)

# 1. Load general config
config <- read_yaml(file.path(app_folder_path, 'general_config.yaml'))

# 2. Source utility scripts
invisible(lapply(list.files(file.path(app_folder_path, 'util_scripts'), pattern = '\\.R$', full.name = TRUE), source))

# 3. Setup paths and load dataset configs
processed_dataset_dir_path <- file.path(dataset_dir_path, config$dataset_paths$processed_dir)
raw_data_dir_path <- file.path(dataset_dir_path, config$dataset_paths$raw_dir)

dataset_config_path <- file.path(dataset_dir_path, config$config_files$dataset)
dataset_config <- read_yaml(dataset_config_path)

samples_config_path <- file.path(dataset_dir_path, config$config_files$samples)
samples_config <- read_yaml(samples_config_path)

# Load global resources
hgnc_complete_set <- fread(file.path(resources_path, config$resource_files$hgnc_complete_set), key = 'symbol')
hgnc_complete_set <- hgnc_complete_set[!(ensembl_gene_id %in% names(table(ensembl_gene_id))[table(ensembl_gene_id) > 1])]
```

## 2. Pre-validation

Verify that the `samples_config.yaml` file has been filled out. If this fails, please open that file, fill in your sample metadata, and re-run.

```{r pre-validation}
prevalidate_setup(dataset_dir_path, config)

samples_info_dt <- rbindlist(samples_config$samples, fill = TRUE, use.names = TRUE)
sample_names <- samples_info_dt$id
```

## 3. Data Extraction

If the extract archive bool var is true and the dataset is still packed in an archive (e.g., `.tar`), this step extracts it into an `extracted/` subfolder. Change to false after extraction is done.

```{r extract-raw-data}
raw_tar_path <- file.path(raw_data_dir_path, archive_name)
extracted_dir_path <- file.path(raw_data_dir_path, 'extracted')

print(paste("Does the file exist?", file.exists(raw_tar_path)))

if (extract_archive) {
  message("Extracting tarball... This may take a while.")
  untar(raw_tar_path, exdir = extracted_dir_path)
  message("Extraction complete.")
} else {
  message("Not extracting")
}
```

## 4. File Mapping & ID Alignment (USER ACTION REQUIRED)

A regex pattern is used to automatically pair the `barcodes`, `features`, and `matrix` files for each sample. An example is shown, but probably needs to be adjusted to new dataset.

::: callout-important
### TODO: Adjust regex pattern and Align Sample IDs

The IDs extracted from the raw file names often do not exactly match the clean IDs you put in `samples_config.yaml` (e.g., supplementary file IDs vs internal IDs). Review the printed tables below and map any discrepancies in the `changes` vector.
:::

```{r map-files}
# --- USER INPUT REQUIRED ---
# Regex pattern to select and pair the three 10x files for each sample
# The given regex pattern works for names like 'GSM6204109_P01_barcodes.tsv.gz'
pattern <- "GSM[0-9]+_([A-Za-z0-9_]+)_(barcodes|features|matrix)\\.(tsv|mtx)\\.gz"

# 10x specific information necessary for mapping to standard
group_indices <- list(sample_id = 1, file_type = 2)
type_mapping <- list(barcodes = "cells", features = "genes", matrix = "matrix")

all_files <- list.files(extracted_dir_path, full.names = TRUE)
file_info <- data.table(path = all_files, file_name = basename(all_files))

matches <- str_match(file_info$file_name, pattern)

# Wrap the data.table assignments in invisible() to suppress Quarto printing
invisible(file_info[, sample_id := matches[, group_indices$sample_id + 1]])
invisible(file_info[, std_type := unlist(type_mapping)[matches[, group_indices$file_type + 1]]])

sample_map <- dcast(file_info[!is.na(sample_id)], sample_id ~ std_type, value.var = "path")

# Display raw IDs vs configured IDs
message("Raw IDs found in file names:")
print(unique(sample_map$sample_id))
message("IDs configured in samples_config.yaml:")
print(sample_names)

# --- USER INPUT REQUIRED ---
# Map any raw IDs (left) to your configured samples_config.yaml IDs (right).
# Leave empty if they already match perfectly.
changes <- c(
  # "raw_file_id_01" = "clean_config_id_01",
  # "KC_08_1" = "CC_02_1" 
)

if (length(changes) > 0) {
  matched_ids <- changes[sample_map$sample_id]
  sample_map$sample_id <- ifelse(is.na(matched_ids), sample_map$sample_id, matched_ids)
}
```

## 5. Matrix Inspection & Configuration

Let's load the first sample's matrix to determine if it contains raw counts, TPM, etc.

```{r matrix-inspection}
# Load the first matrix to check values
test_expmat <- readMM(sample_map$matrix[1])
summary(Matrix::colSums(test_expmat)[1:100]) # View first 100 column sums
max(test_expmat)
```

::: callout-note
### Update Matrix State

Based on the output above, update the `matrix_state` variable. \* If values are strictly integers, it is likely `"UMIcounts"`. \* If column sums equal exactly 10k or 1M, it is likely `"CP10k"` or `"TPM"`.
:::

```{r update-config}
# --- USER INPUT REQUIRED ---
matrix_state <- "UMIcounts" # Update this string

new_expmat_name <- paste0("Exp_data_", matrix_state, ".mtx")
dataset_config$std_names$expmat <- new_expmat_name
write_yaml(dataset_config, dataset_config_path)

message("Updated dataset_config.yaml! Expected matrix name is now: ", new_expmat_name)
```

## 6. Standardization Loop

This automated loop processes every matched 10x sample: loading the sparse matrix, checking gene name formatting (mapping Ensembl to HGNC if necessary), building the `Cells.csv` metadata, and saving the standard outputs.

```{r standardize-loop}
sample_cell_counts <- list()

for (i in seq_len(nrow(sample_map))) {
  sample_name <- sample_map$sample_id[i]
  message(i, '. Standardizing sample: ', sample_name)
  
  # 1. Read files
  expmat <- readMM(sample_map$matrix[i])
  if (!inherits(expmat, "dgCMatrix")) expmat <- as(expmat, "CsparseMatrix")
  
  cell_names <- fread(sample_map$cells[i], header = FALSE)$V1
  genes <- fread(sample_map$genes[i], header = FALSE)
  genes_mapped <- genes$V2
  
  # 2. Check and map genes
  if (any(duplicated(genes$V1))) stop('Ensembl gene ids are not unique')
  if (any(duplicated(genes$V2))) {
    message('Gene symbols are not unique. Mapping using ensembl gene ids...')
    genes_mapped <- map_ensembl_to_symbol_vector(genes$V1, hgnc_complete_set)
  }
  
  if (length(unique(cell_names)) != length(cell_names)) stop('Cell names are not unique')
  
  rownames(expmat) <- genes_mapped
  colnames(expmat) <- cell_names
  
  # Remove NA genes
  valid_genes <- !is.na(genes_mapped)
  expmat <- expmat[valid_genes, ]
  genes_mapped <- genes_mapped[valid_genes]
  
  # 3. Build Cells metadata
  cells <- build_cells(colnames(expmat))
  cells[, sample := sample_name]
  cells[, patient := samples_config$samples[[which(sample_names == sample_name)]]$patient]
  cells[, complexity := matkot::col_nnz(expmat)]
  cells[, source := samples_config$samples[[which(sample_names == sample_name)]]$site]
  
  sample_cell_counts[[sample_name]] <- dim(cells)[1]
  
  # 4. Save
  sample_dir_path <- file.path(processed_dataset_dir_path, sample_name)
  dir.create(sample_dir_path, showWarnings = FALSE)
  
  writeMM(expmat, file = file.path(sample_dir_path, dataset_config$std_names$expmat))
  fwrite(cells, file = file.path(sample_dir_path, dataset_config$std_names$cells))
  fwrite(data.table(genes_mapped), file = file.path(sample_dir_path, dataset_config$std_names$genes), col.names = FALSE)
}
```

## 7. Finalize Pipeline Configuration

Compile the overall `Samples.csv` registry and initialize the clustering/annotation configurations.

```{r finalize-configs}
samples <- build_samples(sample_names) 

cols_to_add <- setdiff(names(samples_info_dt), "id")
invisible(samples[samples_info_dt, (cols_to_add) := mget(paste0("i.", cols_to_add)), on = .(sample = id)])
invisible(samples[, n_cells := unlist(sample_cell_counts)[sample]])

fwrite(samples, file = file.path(processed_dataset_dir_path, dataset_config$std_names$samples))

# Setup selected_config.yaml based on samples
selected_config <- setup_selected_config(samples$sample)
write_yaml(selected_config, file.path(dataset_dir_path, config$config_files$selected))
```

## 8. Post-Validation

Run this final check to ensure the output matches the strict structural requirements of the Shiny App.

```{r post-validation}
#| output: asis

# This will generate HTML warning boxes if something is wrong, 
# or a green success box if you are ready to proceed!
validate_standardization(
  processed_dir_path = processed_dataset_dir_path, 
  sample_names = sample_names, 
  std_names = dataset_config$std_names
)
```
